---
title: "School Experiment"
author: "Andrea Turchetto"
date: "26 January 2017"
output: html_document
---

## School Experiment

This page prensents the results of a study conducted on in two cities (Fresno, Calif. and Yougstown, Ohio) in grades 1, 2, 3 and 4 in several different schools.

The data of the experiment comes from  http://www.stat.columbia.edu/~gelman/bda.course/electric.txt. This analysis is an optional assignment from the DS101x online course: Statistical Thinking for Data Science and Analystics.

The description of the experiment follows:

The classrooms in each grade in each school were paired, with one classroom randomly assigned to the treatment condition (the kids in those classrooms were shown a special educational television show) and the other classroom getting the control condition (no TV show). In each classroom (treated or control), the students were given a "pre-test" at the beginning of the school year and a "post-test" at the end.

Each row of the table shows the data for the two schools in a pair:  the average pre-test score for the students in the treated classroom, the average post-test score for those students, then the average pre- and post-test scores for the students in the control class.  You can ignore the last row of the table. 

The data provided is imported into an R DataFrame df. 
The first row in the txt file states that third and fourth columns present the score for the treated classes and fifth and sixth columns present the results for the control classes. This row is skipped because the title of the columns are in row 2 of the text file. 

```{r }
df = read.table("Experiment.txt", header = TRUE, skip = 1)

## Dimensions of the data frame:
dim(df)
nrow(df)
ncol(df)

## Removing the last row of data - Ignore it as per requirement
df = df[1:(nrow(df)-1),]

## Showing structure of the object
str(df)

## Show first 5 rows of the table
head(df, n=5)

```

We can see that:
Column 0 - City: F (Fresno) / Y (Youngstown)
Column 1 - Grade: 1 to 4
Column 2 - Pretest: average score of initial test for Treated class 
Column 3 - Posttest: average score of final test for Treated class
Column 4 - Pretest.1: average score of initial test for Control class 
Column 5 - Posttest.1: average score of final test for Control class
Column 6 - Supplemment: class is replacement or supplement 

Column 6 is not used for the scope of this analysis

Let's start the exploratory analysis evaluating the statistics of the initial and final tests.

``` {r}
library(dplyr)
# Describe statistics of score tests independently from City or Grade
scores.tot <- summarise(df, count = n(),
                   mean.pre = mean(Pretest),
                   mean.post = mean(Posttest),
                   mean.pre.1 = mean(Pretest.1),
                   mean.post.1 = mean(Posttest.1),
                   median.pre = median(Pretest),
                   median.post = median(Posttest),
                   median.pre.1 = median(Pretest.1),
                   median.post.1 = median(Posttest.1),
                   std.pre = sd(Pretest),
                   std.post = sd(Posttest),
                   std.pre.1 = sd(Pretest.1),
                   std.post.1 = sd(Posttest.1))
scores.tot
```

Let's now describe the statistics for the score based on City and Grade

```{r, library(dplyr)}
# Get statistics grouped by City
by_city <- group_by(df, City)
scores.city <- summarise(by_city,
                   count = n(),
                   mean.pre = mean(Pretest),
                   mean.post = mean(Posttest),
                   mean.pre.1 = mean(Pretest.1),
                   mean.post.1 = mean(Posttest.1),
                   median.pre = median(Pretest),
                   median.post = median(Posttest),
                   median.pre.1 = median(Pretest.1),
                   median.post.1 = median(Posttest.1),
                   std.pre = sd(Pretest),
                   std.post = sd(Posttest),
                   std.pre.1 = sd(Pretest.1),
                   std.post.1 = sd(Posttest.1))
scores.city

# Get statistics grouped by Grade:
by_grade <- group_by(df, Grade)
scores.grade <- summarise(by_grade,
                          count = n(),
                   mean.pre = mean(Pretest),
                   mean.post = mean(Posttest),
                   mean.pre.1 = mean(Pretest.1),
                   mean.post.1 = mean(Posttest.1),
                   median.pre = median(Pretest),
                   median.post = median(Posttest),
                   median.pre.1 = median(Pretest.1),
                   median.post.1 = median(Posttest.1),
                   std.pre = sd(Pretest),
                   std.post = sd(Posttest),
                   std.pre.1 = sd(Pretest.1),
                   std.post.1 = sd(Posttest.1))

subset(scores.grade, select = c("Grade","mean.pre","mean.post","mean.pre.1","mean.post.1"))
```

Let's now plot the scores obtained in the pretest from both the Treated and Control test classes

```{r}
library(ggplot2)
## Factorise Grade variable --> categorical variable
df$Grade <- factor(df$Grade, labels = c("First","Second","Third","Fourth"))

c1 <- ggplot(df, aes(x = Pretest, color = Grade))
c1 + geom_histogram(aes(y=..density..),binwidth = 2) + geom_density()+
  labs(x="Score of Pretest", y = "Density", title = "Probability Distribution of score at Pretest")+theme(legend.position = "top")

c2 <- ggplot(filter(df, Grade == "Second"), aes(Pretest))
c2 + geom_histogram(aes(y = ..density..), binwidth = 5) + geom_density() 

c3 <- ggplot(filter(df, Grade == "Third"), aes(Pretest))
c3 + geom_histogram(aes(y = ..density..), binwidth = 5) + geom_density()

```

Let's now give a look to the increment in score between the initial test and the final test

``` {r}
df = mutate(df, diff_score_treated = Posttest.1-Pretest.1,
           diff_score_control = Posttest-Pretest)

# Get the probability distribution of the differences:
d1 <- ggplot(df, aes(diff_score_control))
d1 + geom_histogram(aes(y = ..density..), binwidth = 5) + geom_density()

d2 <- ggplot(df, aes(diff_score_treated))
d2 + geom_histogram(aes(y = ..density..), binwidth = 5) + geom_density()
```

Let's identify whether the difference between the improvment in test score between treated and control classes is statistically significant. A box plot will do the job.

```{r}
d3 <- ggplot(df, aes( x = Grade, y = diff_score_control))
d3 + geom_boxplot()

## Merge quantitative variables variables diff_score_control and diff_score_treated
df = mutate(df, treated = "T", control = "C")

df1 = select(df, Grade, diff_score_control, control)
colnames(df1)[2:3] <- c("Diff.Score","ClassType")

df2 = select(df, Grade, diff_score_treated, treated)
colnames(df2)[2:3] <- c("Diff.Score","ClassType")

df3 = bind_rows(df1,df2)
rm(df1)
rm(df2)

df3$ClassType = factor(df3$ClassType, levels = c("C","T"), labels = c("Control","Treated"))

str(df3)

d4 <- ggplot(df3, aes(x = ClassType, y = Diff.Score))
d4 + geom_boxplot()
```

From the boxplot above it can be concluded that there is not a statistically significant difference between the improvement in score between the treated class and the control class. 

Let's now check whether the treatment affects the students uniformly over the different ages (grades)

``` {r}
for (grade in 1:length(unique(df3$Grade)))
  {
    d5 <- ggplot(filter(df3, Grade == df3$Grade[grade]), aes(x = ClassType, y = Diff.Score))
    d5 + geom_boxplot()
}
```